---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# sc

The goal of sc is to provide a general common form for complex multi-dimensional data. 

See the [proposal.md](proposal.md).


This package is intended to provide support to the `common form` approach described here. There's very little in the package yet, but see these projects that are informed by this approach. 

* **rbgm** - [Atlantis Box Geometry Model](https://github.com/AustralianAntarcticDivision/rbgm), a "doubly-connected edge-list" form of linked faces and boxes in a spatially-explicit 3D ecosystem model
* **rangl** - [Primitives for Spatial data](https://github.com/r-gris/rangl), a generalization of GIS forms with simple 3D plotting
* **spbabel** - [Translators for R Spatial](https://github.com/mdsumner/spbabel), tools to convert from and to spatial forms, provides the general decomposition framework for branches, used by `rangl`
* **sfct** - [Constrained Triangulation for Simple Features](https://github.com/r-gris/sfct) tools to decompose `simple features` into (non-mesh-indexed) primitives. 

## Example

```{r}
library(sf)
## a MULTIPOLYGON layer
nc = st_read(system.file("shape/nc.shp", package="sf"))
```

The common form is the entity tables, objects, branches, vertices and a link table to allow de-duplication of shared vertices. Currently this de-duplication is done on all coordinate fields, but for most applications it will usually be done only in X-Y. 

```{r}
library(sc)
(bmodel <- BRANCH(nc))
```

Prove that things work by round-tripping to the BRANCH model and onto the old fortify approach for `ggplot2`. 

```{r}
inner_cascade <- function(x) {
  tabnames <- sc:::join_ramp(x)
  tab <- x[[tabnames[1]]]
  for (ni in tabnames[-1L]) tab <- dplyr::inner_join(tab, x[[ni]])
  tab
}

## this just joins everything back together in one big fortify table
library(dplyr)
tab <- bmodel  %>% inner_cascade()

library(ggplot2)
ggplot(tab) + aes(x = x_, y = y_, group = branch_) + 
  geom_polygon(aes(fill = AREA)) +  geom_path(lwd = 2, col = "black") 
```