---
title: "silicate-data-model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{silicate-data-model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(silicate)
```

The silicate package provides tools and data models to work with decompositions of complex data types. The primary examples of use are
with spatial data layers and most silicate workflows are geared to using the underlying structures and topology of spatial data. We can use these for a range of activities: 

* converting data from one format to another
* exploring the underlying relationships with a data set
* developing new tools. 

The functions in silicate are grouped into *models* and *worker verbs*. The models have capitalized names and the worker verbs are prefixed with 'sc'. 

The worker verbs work with all (there are some exceptions) model types and with several external formats, particularly sf, sp, trip, and other external package types.  